Methods for Analyzing Models of\\ Rod-Shaped Bacteria

### 2025-12-03 paper/main.tex
add more details for likelihood approach

% This metric can be used to compare the model output to datapoints obtained from microscopic images.
A common technique to investigate identifiability questions and provide insights into confidence
levels of the obtained results is the
\textit{profile-likelihood}~\cite{Murphy2000,Kreutz2013,Tnsing2023}.
It is able to provide confidence intervals which are invariant under parameter
transformations~\cite{Meeker1995}.
It builds on the so-called \textit{log-likelihood} which propagates uncertainties $\sigma_{ijkl}$ in
the datapoints.
Under the assumption of a constant, known uncertainty, i.e. $\sigma_{ijkl}=\sigma$, we can rewrite
the log-likelihood in terms of $L_\text{LS}$ via
\begin{align}
    L(\theta) &= \sum\limits_{ijkl}\left(\left(\frac{\nu_{ijkl}-\tilde{\nu}_{ijkl}}{\sigma_{ijkl}}\right)^2
    + 2 \log(\sigma_{ijkl})
    +\text{const}\right)\\
    &= \frac{L_\text{LS}(\theta)}{\sigma^2} + \text{const}
\end{align}
where we absorbed the terms $\log(\sigma)$ in the sum into the constant term.
With this simplification, it is important to pick a value for $\sigma$ such that the real
uncertainty is overestimated.
Furthermore, the preceding section (\nameref{subsec:position-extraction-algorithm}) has shown that
the error distribution of the underlying data follows a normal distribution which justifies the core
assumption of this framework.
We conclude that minimizing, the least squares value $L_\text{LS}$ under constant $\sigma$ is
equivalent to minimizing the \textit{log-likelihood} $L$ which is again equivalent to optimizing the
likelihood that the chosen parameters represent our model.

The \textit{profile likelihood} $PL(\theta_i)$ scans a single parameter $\theta_i$ along its axis
while reoptimizing the remaining ones.
We can obtain confidence intervals via
\begin{equation}
    PL(\theta_i) - L(\theta) \leq \Delta_\alpha
\end{equation}
where the value $\Delta_\alpha$ is calculated from the $(1-\alpha)$ quantile of a $\chi^2$
distribution with one degree of freedom~\cite{Tnsing2023}.

>>>>

% This metric can be used to compare the model output to datapoints obtained from microscopic images.
A common technique to investigate identifiability questions and provide insights into confidence
levels of the obtained results is the
\textit{profile-likelihood}~\cite{Murphy2000,Kreutz2013,Tnsing2023}.
It is able to provide confidence intervals which are invariant under parameter
transformations~\cite{Meeker1995}.
It builds on the so-called \textit{log-likelihood} which propagates uncertainties $\sigma_{ijkl}$ in
the datapoints.
Under the assumption of a constant, known uncertainty, i.e. $\sigma_{ijkl}=\sigma$, we can rewrite
the log-likelihood in terms of $L_\text{LS}$ via
\begin{align}
    L(\theta) &= \sum\limits_{ijkl}\left(\left(\frac{\nu_{ijkl}-\tilde{\nu}_{ijkl}}{\sigma_{ijkl}}\right)^2
    + 2 \log(\sigma_{ijkl})
    +\text{const}\right)\\
    &= \frac{L_\text{LS}(\theta)}{\sigma^2} + \text{const}
\end{align}
where we absorbed the terms $\log(\sigma)$ in the sum into the constant term.
With this simplification, it is important to pick a value for $\sigma$ such that the real
uncertainty is overestimated.
Furthermore, the preceding section (\nameref{subsec:position-extraction-algorithm}) has shown that
the error distribution of the underlying data follows a normal distribution which justifies the core
assumption of this framework.
We conclude that minimizing, the least squares value $L_\text{LS}$ under constant $\sigma$ is
equivalent to minimizing the \textit{log-likelihood} $L$ which is again equivalent to maximizing the
likelihood that the chosen parameters represent our model.

The \textit{profile likelihood} $PL(\theta_i)$ makes use of these formalisms and scans a single
parameter $\theta_i$ along its axis while reoptimizing the remaining ones.
We can obtain confidence intervals via
\begin{equation}
    PL(\theta_i) - L(\theta) \leq \Delta_\alpha
    \label{eq:profile-likelihood-confidence-interval}
\end{equation}
where the value $\Delta_\alpha$ is calculated from the $(1-\alpha)$ quantile of a $\chi^2$
distribution with one degree of freedom~\cite{Tnsing2023}.
This means that every parameter $\theta_i$ which satisfies
relation~\eqref{eq:profile-likelihood-confidence-interval} has confidence level $(1-\alpha)$.
In this work we will use confidence intervals $1-\alpha=68\%,90\%,95\%$.
